---
title: "Analysis of COVID-era Apple Mobility Data"
author: "Mikaela Kuan"
date: "3/4/2021"
output: html_document
bibliography: references.bib
params:
  state: "California"
  data: "data/raw_data/applemobilitytrends-2021-03-02.csv"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE)
```

## Setup

```{r set-parameters}
# set the state to work with
state_to_analyze <- params$state

# sanitizing the state name for output files
state_no_spaces <- gsub(state_to_analyze,
                        pattern = " ",
                        replacement = "_")

# set the name of the global data file to use
global_file_to_process <- params$data

# extract core name for output file naming use
core_name <- tools::file_path_sans_ext(basename(global_file_to_process))
```

```{r load-packages-functions}
# load the packages needed
library("dplyr")
library("readr")
library("ggplot2")
library("tidyr")
library("lubridate")
library("devtools")

#load functions needed
source("code/functions/subset_mobility_data_to_state_wide.R")
source("code/functions/subset_mobility_data_to_state_long.R")
source("code/functions/count_up_counties_cities_transportation.R")
```

## Background and Objectives

This rmarkdown file will use `ggplot2` [@ggplot2] and `dplyr` [@dplyr] packages to analyze Apple mobility data from the time before the COVID-19 pandemic through the initial shutdowns and the continuing changes over the course of the year.

The goal will be to create a subset to a particular state, tally up the number of cities and counties in that state with mobility data, and then to make several plots to visualize those data subsets.

The state that this document will analyze is `r state_to_analyze`.

## Data Processing

```{r subset-state}
# subset out to the chosen state using the variables set above
state_subset <- subset_mobility_data_to_state(
  input_file_name = global_file_to_process,
  state_to_subset = state_to_analyze)
```

```{r make-subset-long}
# convert the subsetted state data from wide to long format
long_subset <- subset_data_wide_to_long(
  input_file_name = global_file_to_process,
  state_to_subset = state_to_analyze)
```

```{r tallied-subset}
# tally up the numbers of cities and counties in the state that have mobility
# data available
tallied_subset <- subset_data_count_city_county(
  input_file_name = global_file_to_process,
  state_to_subset = state_to_analyze)
```

## Figures

```{r plot-tallied}
# create a grouped barplot of the tallied data to include in the knitted
# document, and save that figure to the output/figures directory as a png
tallied_plot <- ggplot(data = tallied_subset,
                       aes(x = geo_type,
                           y = n,
                           fill = transportation_type)) +
  geom_col(position = position_dodge()) +
  labs(title = paste("Number of cities and/or counties in",
                     state_to_analyze,
                     "with mobility data"),
       x = "Type of area",
       y = "Number of areas") +
  scale_fill_discrete(name = "Transportation Type",
                      labels = c("Driving", "Transit", "Walking"))

ggsave(plot = tallied_plot,
       filename = paste0("output/figures/",
                         core_name,
                         "_",
                         state_no_spaces,
                         "_tallied_plot.png"))

tallied_plot
```

```{r plot-time-series}
# produce a line plot of relative driving mobility across the state with date
# on the x axis and mean relative mobility on the y axis
timeseries_plot <- long_subset %>%
  filter(transportation_type == "driving") %>%
  group_by(date) %>%
  summarize(mean_mobility = mean(relative_mobility)) %>%
  ggplot(aes(x = lubridate::ymd(date),
             y = mean_mobility)) +
  geom_line() +
  labs(title = paste("Statewide mean relative mobility driving levels in",
                     state_to_analyze,
                     "during COVID"),
       x = "Date",
       y = "Mean relative mobility")

ggsave(plot = timeseries_plot,
       filename = paste0("output/figures/",
                         core_name,
                         "_",
                         state_no_spaces,
                         "_driving_timeseries_plot.png"))

timeseries_plot
```

```{r plot-time-series-full}
# produce a line plot with all three types of transportion with transportation
# on the x-axis and the mean relative mobility on the y axis
driving_timeseries_plot <- long_subset %>%
  filter(transportation_type == "driving") %>%
  group_by(date) %>%
  summarize(mean_mobility_driving = mean(relative_mobility))
transit_timeseries_plot <- long_subset %>%
  filter(transportation_type == "transit") %>%
  group_by(date) %>%
  summarize(mean_mobility_transit = mean(relative_mobility))
walking_timeseries_plot <- long_subset %>%
  filter(transportation_type == "walking") %>%
  group_by(date) %>%
  summarize(mean_mobility_walking = mean(relative_mobility))

# need to create some conditions because some states may not have all the data
# that is need to produce the graph
if (nrow(driving_timeseries_plot) == 0 & nrow(transit_timeseries_plot) == 0) {
  all_timeseries_plot <- ggplot() +
  geom_line(data = walking_timeseries_plot, aes(x = lubridate::ymd(date),
           y = mean_mobility_walking, color = "Walking"), size = 0.71) +
  scale_color_manual(values = c("#FFDB6D", "#FC4E07", "#00AFBB")) +
  theme_linedraw() +
  theme_light() +
  labs(color = "Transportation Type",
       title = paste("Statewide mean relative mobility levels in",
                     state_to_analyze,
                     "during COVID"),
       x = "Date",
       y = "Mean relative mobility")

 ggsave(plot = all_timeseries_plot,
       filename = paste0("output/figures/",
                         core_name,
                         "_",
                         state_no_spaces,
                         "_all_timeseries_plot.png"))
} else if (nrow(driving_timeseries_plot) == 0 &
           nrow(walking_timeseries_plot) == 0) {
  all_timeseries_plot <- ggplot() +
  geom_line(data = transit_timeseries_plot, aes(x = lubridate::ymd(date),
           y = mean_mobility_transit, color = "Transit"), size = 0.71) +
  scale_color_manual(values = c("#FFDB6D", "#FC4E07", "#00AFBB")) +
  theme_linedraw() +
  theme_light() +
  labs(color = "Transportation Type",
       title = paste("Statewide mean relative mobility levels in",
                     state_to_analyze,
                     "during COVID"),
       x = "Date",
       y = "Mean relative mobility")

 ggsave(plot = all_timeseries_plot,
       filename = paste0("output/figures/",
                         core_name,
                         "_",
                         state_no_spaces,
                         "_all_timeseries_plot.png"))
} else if (nrow(transit_timeseries_plot) == 0 &
           nrow(walking_timeseries_plot) == 0) {
  all_timeseries_plot <- ggplot() +
  geom_line(data = driving_timeseries_plot,
            aes(x = lubridate::ymd(date), y = mean_mobility_driving,
                color = "Driving"), size = 0.71) +
  scale_color_manual(values = c("#FFDB6D", "#FC4E07", "#00AFBB")) +
  theme_linedraw() +
  theme_light() +
  labs(color = "Transportation Type",
       title = paste("Statewide mean relative mobility levels in",
                     state_to_analyze,
                     "during COVID"),
       x = "Date",
       y = "Mean relative mobility")

 ggsave(plot = all_timeseries_plot,
       filename = paste0("output/figures/",
                         core_name,
                         "_",
                         state_no_spaces,
                         "_all_timeseries_plot.png"))
} else if (nrow(driving_timeseries_plot) == 0) {
  all_timeseries_plot <- ggplot() +
  geom_line(data = transit_timeseries_plot, aes(x = lubridate::ymd(date),
           y = mean_mobility_transit, color = "Transit"), size = 0.71) +
  geom_line(data = walking_timeseries_plot, aes(x = lubridate::ymd(date),
           y = mean_mobility_walking, color = "Walking"), size = 0.71) +
  scale_color_manual(values = c("#FFDB6D", "#FC4E07", "#00AFBB")) +
  theme_linedraw() +
  theme_light() +
  labs(color = "Transportation Type",
       title = paste("Statewide mean relative mobility levels in",
                     state_to_analyze,
                     "during COVID"),
       x = "Date",
       y = "Mean relative mobility")

 ggsave(plot = all_timeseries_plot,
       filename = paste0("output/figures/",
                         core_name,
                         "_",
                         state_no_spaces,
                         "_all_timeseries_plot.png"))
} else if (nrow(transit_timeseries_plot) == 0) {
  all_timeseries_plot <- ggplot() +
  geom_line(data = driving_timeseries_plot,
            aes(x = lubridate::ymd(date), y = mean_mobility_driving,
                color = "Driving"), size = 0.71) +
  geom_line(data = walking_timeseries_plot, aes(x = lubridate::ymd(date),
           y = mean_mobility_walking, color = "Walking"), size = 0.71) +
  scale_color_manual(values = c("#FFDB6D", "#FC4E07", "#00AFBB")) +
  theme_linedraw() +
  theme_light() +
  labs(color = "Transportation Type",
       title = paste("Statewide mean relative mobility levels in",
                     state_to_analyze,
                     "during COVID"),
       x = "Date",
       y = "Mean relative mobility")

 ggsave(plot = all_timeseries_plot,
       filename = paste0("output/figures/",
                         core_name,
                         "_",
                         state_no_spaces,
                         "_all_timeseries_plot.png"))
} else if (nrow(walking_timeseries_plot) == 0) {
  all_timeseries_plot <- ggplot() +
  geom_line(data = driving_timeseries_plot,
            aes(x = lubridate::ymd(date), y = mean_mobility_driving,
                color = "Driving"), size = 0.71) +
  geom_line(data = transit_timeseries_plot, aes(x = lubridate::ymd(date),
           y = mean_mobility_transit, color = "Transit"), size = 0.71) +
  scale_color_manual(values = c("#FFDB6D", "#FC4E07", "#00AFBB")) +
  theme_linedraw() +
  theme_light() +
  labs(color = "Transportation Type",
       title = paste("Statewide mean relative mobility levels in",
                     state_to_analyze,
                     "during COVID"),
       x = "Date",
       y = "Mean relative mobility")

 ggsave(plot = all_timeseries_plot,
       filename = paste0("output/figures/",
                         core_name,
                         "_",
                         state_no_spaces,
                         "_all_timeseries_plot.png"))
} else {
  all_timeseries_plot <- ggplot() +
  geom_line(data = driving_timeseries_plot,
            aes(x = lubridate::ymd(date), y = mean_mobility_driving,
                color = "Driving"), size = 0.71) +
  geom_line(data = transit_timeseries_plot, aes(x = lubridate::ymd(date),
           y = mean_mobility_transit, color = "Transit"), size = 0.71) +
  geom_line(data = walking_timeseries_plot, aes(x = lubridate::ymd(date),
           y = mean_mobility_walking, color = "Walking"), size = 0.71) +
  scale_color_manual(values = c("#FFDB6D", "#FC4E07", "#00AFBB")) +
  theme_linedraw() +
  theme_light() +
  labs(color = "Transportation Type",
       title = paste("Statewide mean relative mobility levels in",
                     state_to_analyze,
                     "during COVID"),
       x = "Date",
       y = "Mean relative mobility")

 ggsave(plot = all_timeseries_plot,
       filename = paste0("output/figures/",
                         core_name,
                         "_",
                         state_no_spaces,
                         "_all_timeseries_plot.png"))
}

all_timeseries_plot
```

# Session Info

```{r session-info}
devtools::session_info()
```

## Sources Cited
